### 2023-06-29 journal/algorithm-GC.tex
created directory for eventual journal version

Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode2GC}.
New or modified code appears in blue.
The \op{Propagate} and \op{GetEnqueue} routines are unchanged.
A few lines have been added to \op{FindResponse} to update the \var{last} array
to ensure that it stores the value described above.
Minor modifications have also
been made to \op{Enqueue}, \op{Dequeue}, \op{CreateBlock}, \op{Refresh} and \op{Append}
to accommodate  the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
In addition, the second half of the \op{Dequeue} routine is now in a
separate routine called \op{CompleteDeq} so that it can also be used by other processes
helping to complete the operation.
The \op{Refresh} routine no longer needs to set the \fld{super} field of blocks since
that field has been removed.
The \op{IndexDequeue} routine, which must trace the location of a dequeue along
a path from its leaf to the root has a minor modification to search the \fld{blocks} RBT at 
each level instead of using the \fld{super} field.

@@

To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue using the \var{last} array.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
helped to complete by line \ref{invokeHelp},
it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
along the path from \var{v} to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).

@@

In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
with index~$i$.
A search for this block may sometimes not find it, if it has already been discarded
by another process's GC.
As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
the enqueue can simply terminate because the fact that the block is gone means that another
process has helped the enqueue reach the root of the ordering tree.
Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
written in the \fld{response} field of the 
leaf block that represents the dequeue and terminate, since some other process
will have written the \fld{response} there before discarding the needed block.
We do not explicitly write this early termination every time we do a lookup in an RBT.
There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
on line \ref{SBlookup1} or \ref{SBlookup2} of
\op{SplitBlock} because the required block has been discarded, 
we continue doing GC, since we do not a GC phase on one node
to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
that were needed.
Line \ref{substitute} says what to do in this case.

>>>>

Pseudocode for the space-bounded implementation appears in Figures \ref{pseudocode1GC} and \ref{pseudocode2GC}.
New or modified code appears in blue.
The \op{Propagate} and \op{GetEnqueue} routines are unchanged.
A few lines have been added to \op{FindResponse} to update the \var{last} array
to ensure that it stores the value described above.
Minor modifications have also
been made to \op{Enqueue}, \op{Dequeue}, \op{CreateBlock}, \op{Refresh} and \op{Append}
to accommodate  the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
In addition, the second half of the \op{Dequeue} routine is now in a
separate routine called \op{CompleteDeq} so that it can also be used by other processes
helping to complete the operation.
The \op{Refresh} routine no longer needs to set the \fld{super} field of blocks since
that field has been removed.
The \op{IndexDequeue} routine, which must trace the location of a dequeue along
a path from its leaf to the root has a minor modification to search the \fld{blocks} RBT at 
each level instead of using the \fld{super} field.

@@

To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
uses the \var{last} array to
find the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue.  By the FIFO property of queues, all enqueues in blocks before $B_{root}$ are either
dequeued or will be dequeued by a dequeue that is currently in progress.  Once those pending dequeues have been
helped to complete by line \ref{invokeHelp},
it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.\footnote{If we used the more conservative approach of discarding blocks whose indices 
are smaller than the \emph{minimum} entry of \var{last} instead of the maximum, helping would be unnecessary, but then one slow process could prevent GC from discarding any blocks, so the space would not be bounded.}
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last block in \var{v} that is a subblock
of $B_{root}$ (or any older block in the root, in case $B_{root}$ has no subblocks in \var{v}).
As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
along the path from \var{v} to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
(since a GC phase on $\var{v}'$ determined that all blocks older than that are safe to discard anyway).

@@

In the code, we use $\var{v}.\fld{blocks}[i]$ to refer to the block in the RBT stored in $\var{v}.\fld{blocks}$
with index~$i$.
A search for this block may sometimes not find it, if it has already been discarded
by another process's GC phase.
As mentioned in Section \ref{reducing}, if this happens to an enqueue operation,
the enqueue can simply terminate because the fact that the block is gone means that another
process has helped the enqueue reach the root of the ordering tree.
Similarly, if  a dequeue operation performs a failed search on a RBT, the dequeue can return the value 
written in the \fld{response} field of the 
leaf block that represents the dequeue and terminate, since some other process
will have written the \fld{response} there before discarding the needed block.
We do not explicitly write this early termination in the pseudocode every time we do a lookup in an RBT.
There is one exception to this rule:  if an RBT lookup for block $B$ returns \nl\
on line \ref{SBlookup1} or \ref{SBlookup2} of
\op{SplitBlock} because the required block has been discarded, 
we continue doing GC, since we do not want a GC phase on one node
to be prevented from cleaning up its RBT because a GC phase on a different node threw away some blocks
that were needed.
Line \ref{substitute} says what to do in this case.

### 2023-06-29 journal/algorithm.tex
created directory for eventual journal version

Once a dequeue is appended to a block of the process's leaf and propagated to the root,
the \opemph{IndexDequeue} routine finds the dequeue's location in the root.
More precisely, \opa{IndexDequeue}{v, b, i}
computes the block in the root and the rank
within that block  of the \var{i}th dequeue of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may be off by one from the superblock's true index.
The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the dequeue's 
rank within the superblock's sequence of dequeues, using  (\ref{defSeqs}).

@@

To compute the response of the $i$th \op{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If not, line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
the \var{e}th enqueue.
Since the enqueue is linearized before the dequeue, $b_e\leq b$.  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

>>>>

Once a dequeue is appended to a block of the process's leaf and propagated to the root,
the \opemph{IndexDequeue} routine finds the dequeue's location in the root.
More precisely, \opa{IndexDequeue}{v, b, i}
computes the block in the root and the rank
within that block  of the \var{i}th dequeue of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
Lines \ref{idsup1}--\ref{idsup3} compute the location of $B$'s superblock in \var{v}'s parent, taking into account the fact that $B.\fld{super}$ may differ from the superblock's true index by one.
The arithmetic in lines \ref{computeISuperStart}--\ref{computeISuperEnd} compute the dequeue's 
rank within the superblock's sequence of dequeues, using  (\ref{defSeqs}).

@@

To compute the response of the $i$th \op{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If not, line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
the \var{e}th enqueue.
Since the enqueue is linearized before the dequeue, $b_e\leq b$.  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} traces down through the tree to find the required enqueue in a leaf.

### 2023-06-29 journal/correctness-GC.tex
created directory for eventual journal version

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{casGC} 
installs $B'$ in $\var{v}.\fld{blocks}$.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
and therefore \var{v} contains $op$.

@@

We say a block is \emph{finished} if 
\begin{itemize}
\item
it has been propagated to the root,
\item
the value of each enqueue contained in the block has either been returned by a dequeue
or written in the \fld{response} field of a dequeue, and
\item
each dequeue contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
\end{itemize}
Intuitively, once a block is finished, operations no longer need to pass through the block
to compute responses to operations.
The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.

@@

We show that every step preserves the invariant.  
We need only consider a step $st$ that modifies a node's RBT.
The minimum \fld{index} of \var{v}'s RBT can only change when \var{v}'s RBT changes, either at line 
\ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
In either case, the step $st$ changes $\var{v}.\fld{blocks}$ is changed
from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
(In the case of a leaf \var{v}, this is true because only the process that owns the leaf ever
writes to $\var{v}.\fld{blocks}$.)
If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
then $T'$ is obtained by adding a new block to $T$, so by \Cref{RBTupdates} the minimum \fld{index}
is unchanged and the invariant is trivially preserved.
So consider the case where $A$ performs GC.
We must show that any block of \var{v} whose index is less than or equal to the
minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
a block with a larger index), it suffices to show that all blocks that were ever added
to \var{v}'s RBT with \fld{index} at most $s$ are finished.

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations of $\var{root}.\fld{blocks}[1..m]$.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates, the $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.

@@

Otherwise, \op{SplitBlock}($x$) returns the block $B$ at line \ref{substitute}.
By the induction hypothesis, the block $B_p$ computed at line \ref{SBrecurse}
(and all earlier blocks of $x$) are finished when $st$ occurs.
By \Cref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
is also finished when $st$ occurs.
\renewcommand{\qedsymbol}{$\diamondsuit$}
\end{proof}

@@

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in its \fld{response} field.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by \Cref{lem::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \op{response} field.

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
$A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
must still be in progress at $C$; otherwise the process that dequeued it would have
set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
contradicting the definition of $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

@@

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation takes is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once as follows.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\index>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.

Each helping phase takes $O(p)$ steps to read the \var{last} array, and
$O(p \log p \log(p+q_{max}))$ steps to \op{Help},
and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on \var{v}.
We can amortize these steps over the operations that appear in 
$\var{v}.\fld{blocks}[(r-1)p^2+1..p^2]$.
By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$.
\end{proof}

The implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still only does two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.

>>>>

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$
that is installed in $\var{v}.\fld{blocks}$
by $R$'s \op{CAS} at line \ref{casGC}.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
and therefore \var{v} contains $op$.

@@

We say a block is \emph{finished} if 
\begin{itemize}
\item
it has been propagated to the root,
\item
the value of each enqueue contained in the block has either been returned by a dequeue
or written in the \fld{response} field of a dequeue, and
\item
each dequeue contained in the block has terminated or some process has written to the \fld{response} field in the leaf block that represents it.
\end{itemize}
Intuitively, once a block is finished, operations no longer need  the block
to compute responses to operations.
The following is an immediate consequence of the definition of finished and what it means for an operation to be contained in a block.

@@

We show that every step preserves the invariant.  
We need only consider a step $st$ that modifies a node's RBT.
The minimum \fld{index} of \var{v}'s RBT can only change when \var{v}'s RBT changes, either at line 
\ref{appendLeafGC} (if \var{v} is a leaf) or at line \ref{casGC} (if \var{v} is an internal node).
In either case, the step $st$ changes $\var{v}.\fld{blocks}$ 
from $T$ to $T'$, where $T'$ is obtained by  a call $A$ to \op{AddBlock}(\var{v}, $T$, $B$).
(In the case of a leaf \var{v}, this is true because only the process that owns the leaf ever
writes to $\var{v}.\fld{blocks}$.)
If $A$ does not do GC (lines \ref{GCstart}--\ref{GCend}), 
then $T'$ is obtained by adding a new block to $T$, so by \Cref{RBTupdates} the minimum \fld{index}
is unchanged and the invariant is trivially preserved.
So consider the case where $A$ performs GC.
We must show that any block of \var{v} whose index is less than or equal to the
minimum \fld{index} in $T'$ is finished when $T'$ is installed in $\var{v}.\fld{blocks}$.
Since $T'$ is obtained by discarding all blocks with \fld{index} at most $s$ (and adding
a block with a larger index), it suffices to show that all blocks that were ever added
to \var{v}'s RBT with \fld{index} at most $s$ are finished.

@@

For the base case, suppose $x$ is the root.
\op{SplitBlock} finds the maximum value $m$ in \var{last}, which is the index of some block
that contains an operation that is either a null dequeue or an enqueue whose value is the response for a dequeue that has been propagated to the root (since these are the only ways that an entry of \var{last} can be set to $m$).
By the FIFO property of queues, the values enqueued by enqueues in $\var{root}.\fld{blocks}[1..m-1]$
are all dequeued by operations that have already been installed in $\var{root}.\fld{blocks}$ before
the end of the \op{SplitBlock}.
Between the termination of $A$'s call to \op{SplitBlock} at line \ref{callSplitBlock}
and the \op{CAS} step $st$ after $A$ terminates,  $A$ helps all pending dequeues at line \ref{invokeHelp}.
Thus, after this helping (and before step $st$),
$\var{root}.\fld{blocks}[1..m-1]$ are all finished blocks, so the claim is true.

@@

Otherwise, \op{SplitBlock}($x$) returns the block $B$ at line \ref{substitute}.
By the induction hypothesis, the block $B_p$ computed at line \ref{SBrecurse}
(and all earlier blocks of $x$) are finished when $st$ occurs.
By \Cref{subblocksFinished}, the block $B$ in $x$ indexed by $B_p.\eleft$ or $B_p.\eright$
is also finished when $st$ occurs.  This completes the proof of \Cref{splitfinished}.
\renewcommand{\qedsymbol}{$\diamondsuit$}
\end{proof}

@@

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in its \fld{response} field.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by \Cref{lem::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by \Cref{lem::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
must still be in progress at~$C$; otherwise the process that dequeued it would have
set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
contradicting the definition of $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

@@

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from \Cref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.

We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.

Each helping phase takes $O(p)$ steps to read the \var{last} array, and
$O(p \log p \log(p+q_{max}))$ steps in \op{Help},
and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on \var{v}.
We can amortize these steps over the operations that appear in 
$\var{v}.\fld{blocks}[(r-1)G+1..rG]$.
By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$.
\end{proof}

The implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.

### 2023-06-29 journal/correctness-details.tex
created directory for eventual journal version

If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1]+1$ and
$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.

@@

\superRelationRes*
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

@@

\indexDequeueRes*
\begin{proof}
We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]$.

>>>>

If $\var{v}$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{enq}}+1$ and
$B.\fld{sum\sub{deq}}$ is set to $\var{v}.\fld{blocks}[i-1].\fld{sum\sub{deq}}$, the claim follows from the hypothesis.
The proof for line~\ref{deqNew}, where $B$ has a single dequeue, is similar.

@@

\superRelationRes*
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} or \ref{createEndRight} obtains a value greater than $b$.
By \Cref{lem::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by \Cref{lem::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by \Cref{nonDecreasingHead}, it follows that $B.super\leq s$.

@@

\indexDequeueRes*
\begin{proof}
We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.
